Place of death in NRS COVID-19 deaths

### 2021-03-10 covid_nrs_place_of_death.Rmd
added a few more plots, and split off sections for paper to a separate file

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
```

@@

@@

@@

# theme_set(theme_minimal(base_size = 16) +
theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri")
            ))

@@

walk(
  .x = extensions,
  .f = ~ggsave(filename = paste0(filename, .x), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (.x==".pdf") cairo_pdf else NULL)
)

@@

@@

```{r, fig.height=12, fig.width=8}
# TODO: work out the size of these plots
figure_map_ratio_home_deaths_2020_to_historical_la <-
  total_deaths_la %>%
  filter(
    year == 2020 & place_of_death == "Home & other non-institution"
  ) %>%
  mutate(percent_increase_home_death = ratio_annual_deaths_to_historical_mean - 1) %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  # ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
  ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x="", y="",
       title = str_wrap("Ratio of home deaths in 2020 to historical mean 2015-2019, by Local Authority", width = 50),
       subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       # fill = "Ratio"
       fill = "Home death increase"
       ) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_la, width = 9, height = 12, dpi = 300, units = "in")

@@

@@

```{r, fig.height=12, fig.width=8}
# TODO: work out the size of these plots
figure_map_ratio_home_deaths_2020_to_historical_hb <-
  total_deaths_hb %>%
  filter(
    year == 2020 & place_of_death == "Home & other non-institution"
  ) %>%
  mutate(percent_increase_home_death = ratio_annual_deaths_to_historical_mean - 1) %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  # ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
  ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(labels = function(x) scales::percent(x, accuracy = 1)) +
  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
  labs(x="", y="",
       title = str_wrap("Ratio of home deaths in 2020 to historical mean 2015-2019, by Health Board", width=50),
       subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       # fill = "Ratio"
       fill = "Home death increase"
       ) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_hb, width = 9, height = 12, dpi = 300, units = "in")

@@

@@

```{r, fig.height=12, fig.width=8}
figure_map_per_capita_covid_cases_2020_la <-
  weekly_covid_cases_la %>%
  filter(year == 2020 & week_number == 53) %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=crude_rate_positive_per100k), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(option = "B") +
  # theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x="", y="",
       title = str_wrap("Cumulative cases of COVID-19 in 2020 per 100,000 population, by Local Authority",width=50),
       subtitle = "",
       caption = paste0(c(areas_relatively_higher_number_cases, source_phs_text), collapse="\n"),
       fill = str_wrap("COVID-19 cases in 2020 per 100,000 population", width = 20)
       ) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_la, width = 9, height = 12, dpi = 300, units = "in")

@@

@@

```{r, fig.height=12, fig.width=8}
figure_map_per_capita_covid_cases_2020_hb <-
  weekly_covid_cases_hb %>%
  filter(year == 2020 & week_number == 53) %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=crude_rate_positive_per100k), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(option = "B") +
  # theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
  labs(x="", y="",
       title = str_wrap("Cumulative cases of COVID-19 in 2020 per 100,000 population, by Health Board",width=50),
       subtitle = "",
       caption = paste0(c(areas_relatively_higher_number_cases, source_phs_text), collapse="\n"),
       fill = str_wrap("COVID-19 cases in 2020 per 100,000 population", width = 20)
       ) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_hb, width = 9, height = 12, dpi = 300, units = "in")

@@

@@

>>>>

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
```

@@

## Weekly deaths including past weekly deaths

For this dataset, we simply `bind_rows` weekly deaths by year and place of death with past weekly deaths by year and location. Note that past data for years with 52 weeks includes an entry of 0 deaths for week 53, so we'll remove them, otherwise the date computation is thrown off!

```{r}
weekly_deaths_current_and_past <-
  bind_rows(
    weekly_deaths_overall %>%
    filter(sex=="all" & age=="all" & place_of_death!="All") %>%  # keep only aggregated values over sex & age
    select(year, week_number, week_number_run_over, place_of_death, matches("deaths"), -deaths_nrs_past_average_all_causes)
    ,
    past_deaths_overall %>%
      rename(deaths_all_causes = number_of_deaths) %>%
      filter(place_of_death!="All" & year!="average") %>%  # remove aggregated entries
      mutate(year = as.integer(year)) %>%
      filter(!(year %in% 2016:2019 & week_number==53))  # remove week 53 entries from years 2016-2019
  ) %>%
  mutate(date_w_c = compute_start_date_from_week_number(week_number = week_number, year_number = year))
```

@@

## Including non-COVID deaths

Note: non-covid deaths are defined as all-cause deaths minus covid-related deaths. Otherwise the number of non-covid deaths only includes non-covid deaths once covid-related deaths started being reported in 2020 (week 12).

```{r}
proportions_of_current_non_covid_deaths_by_place <-
  total_deaths_overall %>%
  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related) %>%
  select(year, place_of_death, number_of_deaths = deaths_non_covid) %>%
  filter(place_of_death != "All") %>%
  group_by(year) %>%
  mutate(
    total_deaths = sum(number_of_deaths)
  ) %>%
  ungroup %>%
  mutate(proportion_of_total = number_of_deaths / total_deaths)

merged_proportions_of_non_covid_deaths_by_place <-
  bind_rows(
    proportions_of_historical_deaths_by_place,  # all historical deaths are non-covid!
    proportions_of_current_non_covid_deaths_by_place
  )
```

@@

theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0)),
                  text=element_text(family="Calibri")
            ))

@@

walk(
  .x = extensions,
  .f = ~function(ext) {
    if (tolower(ext)==".pdf") device <- cairo_pdf

  ggsave(filename = paste0(filename, .x), plot = plot, width = width, height = height, units = units, dpi = dpi, device = if (.x==".pdf") cairo_pdf else NULL)
  }
)

@@

text_annotation_cunulative_all_cause_deaths <- "Hospital all-cause deaths were within the historical range; home & care home deaths exceeded historical ranges."

# TODO: these figures over- and underestimate yearly cumulative deaths (because they use weekly min/max, as opposed to min/max of cumulative deaths)

@@

```{r, fig.height=12, fig.width=8}
# TODO: work out the size of these plots
figure_map_ratio_home_deaths_2020_to_historical_la <-
  total_deaths_la %>%
  filter(
    year == 2020 & place_of_death == "Home & other non-institution"
  ) %>%
  mutate(percent_increase_home_death = ratio_annual_deaths_to_historical_mean - 1) %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(labels = function(x) scales::percent(x, accuracy = 1)) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x="", y="",
       title = str_wrap("Increased home deaths in 2020 relative to historical mean 2015-2019, by Local Authority", width=50),
       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       caption = paste0(c(source_nrs_text), collapse="\n"),
       fill = "Home death\nincrease"
       ) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_la, width = 12, height = 14, dpi = 300, units = "cm")

@@

#### Grayscale version of the same

```{r}
figure_map_ratio_home_deaths_2020_to_historical_la_bw <-
  figure_map_ratio_home_deaths_2020_to_historical_la +
  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) scales::percent(x, accuracy = 1)) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_la_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_la_bw, width = 12, height = 14, dpi = 300, units = "cm")
```

@@

```{r, fig.height=12, fig.width=8}
figure_map_ratio_home_deaths_2020_to_historical_hb <-
  total_deaths_hb %>%
  filter(
    year == 2020 & place_of_death == "Home & other non-institution"
  ) %>%
  mutate(percent_increase_home_death = ratio_annual_deaths_to_historical_mean - 1) %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  # ggplot(aes(geometry=geometry, fill=ratio_annual_deaths_to_historical_mean), colour="grey10") +
  ggplot(aes(geometry=geometry, fill=percent_increase_home_death), colour="grey10") +
  geom_sf(colour="grey10") +
  # scale_fill_viridis_b(limits = c(minimum_ratio_between_hb_and_la_home_deaths-1, maximum_ratio_between_hb_and_la_home_deaths-1), labels = function(x) scales::percent(x, accuracy = 1)) +
  scale_fill_viridis_b(labels = function(x) scales::percent(x, accuracy = 1)) +
  theme(
    axis.text = element_blank(), 
    panel.grid = element_blank(), 
    aspect.ratio = 1.4,  # needed for Scotland HB plot
    plot.margin = margin(5,5,5,5),
    # legend.position = c(1.2,0.4)  # this was manually tweaked until just right  # problem - edge gets cut
    ) +
  labs(x="", y="",
       title = str_wrap("Increased home deaths in 2020 relative to historical mean 2015-2019, by Health Board", width=50),
       # subtitle = str_wrap("A ratio of 1 means there was no increase; A ratio of 1.6 means there was a 60% increase of home deaths in 2020 over the mean 2015-2019.", width = 80),
       # caption = paste0(c(areas_relatively_higher_number_home_deaths_text, home_deaths_ratio_text, source_nrs_text), collapse="\n"),
       caption = paste0(c(source_nrs_text), collapse="\n"),
       # fill = "Ratio"
       fill = "Home death\nincrease"
       ) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_hb, width = 12, height = 14, dpi = 300, units = "cm")

@@

#### Grayscale version of the same

```{r}
figure_map_ratio_home_deaths_2020_to_historical_hb_bw <-
  figure_map_ratio_home_deaths_2020_to_historical_hb +
  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) scales::percent(x, accuracy = 1)) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_ratio_home_deaths_2020_to_historical_hb_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_ratio_home_deaths_2020_to_historical_hb_bw, width = 12, height = 14, dpi = 300, units = "cm")
```

@@

```{r, fig.height=12, fig.width=8}
figure_map_per_capita_covid_cases_2020_la <-
  weekly_covid_cases_la %>%
  filter(year == 2020 & week_number == 53) %>%
  left_join(shapefile_la, by=c("ref_area"="LAD20NM")) %>%
  ggplot(aes(geometry=geometry, fill=crude_rate_positive_per100k), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
  theme(axis.text = element_blank(), panel.grid = element_blank()) +
  labs(x="", y="",
       title = str_wrap("Total COVID-19 cases in 2020 per 100,000 population, by Local Authority",width=50),
       subtitle = "",
       caption = paste0(c(source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_la"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_la, width = 12, height = 14, dpi = 300, units = "cm")

@@

#### Grayscale version of the same

```{r}
figure_map_per_capita_covid_cases_2020_la_bw <-
  figure_map_per_capita_covid_cases_2020_la +
  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) format(x, big.mark=",")) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_la_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_la_bw, width = 12, height = 14, dpi = 300, units = "cm")
```

@@

```{r, fig.height=12, fig.width=8}
figure_map_per_capita_covid_cases_2020_hb <-
  weekly_covid_cases_hb %>%
  filter(year == 2020 & week_number == 53) %>%
  left_join(shapefile_hb_uk_wide, by=c("ref_area"="name")) %>%
  ggplot(aes(geometry=geometry, fill=crude_rate_positive_per100k), colour="grey10") +
  geom_sf(colour="grey10") +
  scale_fill_viridis_b(option = "B", labels = function(x) format(x, big.mark=",")) +
  theme(axis.text = element_blank(), panel.grid = element_blank(), aspect.ratio = 1.4) +
  labs(x="", y="",
       title = str_wrap("Total cases of COVID-19 in 2020 per 100,000 population, by Health Board",width=50),
       subtitle = "",
       caption = paste0(c(source_phs_text), collapse="\n"),
       fill = str_wrap("Cases per 100,000 population", width = 15)
       ) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_hb, width = 12, height = 14, dpi = 300, units = "cm")

@@

#### Grayscale version of the same

```{r}
figure_map_per_capita_covid_cases_2020_hb_bw <-
  figure_map_per_capita_covid_cases_2020_hb +
  scale_fill_gradient(low = "grey10", high = "grey75", labels = function(x) format(x, big.mark=",")) +
  NULL

save_output_file(filename = paste0(dir_outputs,"/figure_map_per_capita_covid_cases_2020_hb_bw"), extensions = c(".pdf",".svg",".png"), plot = figure_map_per_capita_covid_cases_2020_hb_bw, width = 12, height = 14, dpi = 300, units = "cm")
```

@@

# Save most recent workspace for further analyses

```{r}
save.image(file = "./workspace.RData", safe = TRUE)
```

# Print session info

```{r}
sessionInfo()
```
